Abstract Data Type Inference

### 2015-03-24 draft.tex
Some algebra.
Added references to nature.
Added notion of “weight”.
Better Section 2.
Better organization.
Better titles.
Better definition of weight, includes duplicates.
Soft description of the formulas.
Split algorithm section in two.
Added a figure.
started wqo proof
Closure under unmatched-removal.
More on nature.
Updated intro.

\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

@@

\authorinfo{Michael Emmi}
           {IMDEA Software Institute}
           {michael.emmi@imdea.org}
\authorinfo{Constantin Enea\and Jad Hamza}
           {Université Paris Diderot}
           {\{cenea,jhamza\}@univ-paris-diderot.fr}

@@

  Checking that efficient implementations of concurrent objects adhere to their
  abstract data types (ADTs) can be made tractable when ADTs are given by
  symbolic, logical representations. Obtaining these symbolic representations,
  however, is a complex task requiring rare expertise. In practice, concurrent
  objects are specified indirectly by imperative reference implementations.

  In this work we demonstrate that effective symbolic ADT representations can
  be automatically generated from the executions of reference implementations.
  Our approach exploits key features of naturally-occurring ADTs: violations
  can be decomposed into a small set of representative patterns, and these
  patterns manifest in executions with few operations. By identifying certain
  algebraic properties of naturally-occurring ADTs, and exhaustively sampling
  executions up to a small number of operations, we generate concise symbolic
  ADT representations which are complete in practice, enabling the application
  of efficient symbolic verification algorithms without the burden of manual
  specification.

@@

\input intro
\input inference
\input algebra
\input algorithm
\input matching
\input empirical
\input related

>>>>

\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\img}{img}
\DeclareMathOperator{\width}{width}
\DeclareMathOperator{\weight}{weight}
\DeclareMathOperator{\freq}{freq}

\newcommand{\set}[1]{{\{{#1}\}}}
\newcommand{\mset}[1]{{\{\!\{ #1 \}\!\}}}
\newcommand{\tup}[1]{{\left\langle{#1}\right\rangle}}
\renewcommand{\implies}{\Rightarrow}

@@

\authorinfo{}{}{}

% \authorinfo{Michael Emmi}
%            {IMDEA Software Institute}
%            {michael.emmi@imdea.org}
% \authorinfo{Constantin Enea}
%            {Université Paris Diderot}
%            {cenea@univ-paris-diderot.fr}

@@

  Formal specifications are a vital component to the automated verification of
  software systems. In the case of efficient implementations of concurrent
  objects like atomic registers, queues, and locks, checking adherence to their
  abstract data types (ADTs) is a difficult problem, yet can be made tractable
  when the ADT specifications are given by symbolic, logical representations.
  Obtaining these symbolic representations, however, is a complex task
  requiring rare expertise. In practice, concurrent objects are specified
  indirectly by imperative reference implementations.

  In this work we demonstrate that effective symbolic ADT representations can
  be automatically generated from the executions of reference implementations.
  Our approach exploits two key features of naturally-occurring ADTs: violations
  can be decomposed into a small set of representative patterns, and these
  patterns manifest in executions with few operations. By identifying certain
  algebraic properties of naturally-occurring ADTs, and exhaustively sampling
  executions up to a small number of operations, we generate concise symbolic
  ADT representations which are complete in practice, enabling the application
  of efficient symbolic verification algorithms without the burden of manual
  specification. Furthermore, the concise ADT violation patterns we generate
  are human-readable, and can serve as useful, formal documentation.

@@

\input intro
\input histories
\input inference
\input patterns
\input formula
\input matching
\input nature
\input discussion
\input related

### 2015-03-24 intro.tex
Updated intro.

The fundamental challenge is in identifying the algebraic properties of ADTs
which allow us to characterize an infinite set of violating executions with a
finite set of patterns. This characterization is non-trivial since an execution
with more operations than a given violating execution is not necessarily a
violation itself. For instance, an execution which contains only a single {\tt
pop} operation returning the value {\tt 1} is a violation to the atomic stack
ADT, whereas an execution containing an additional {\tt push(1)} operation,
overlapping in time with the {\tt pop}, is not. Further complication arises
from the infinite set of possible data values, i.e.,~method argument and return
values. Our patterns must be sensitive to the \emph{relation} among data values
without being sensitive to the data values themselves. For instance, a
sequential execution in which {\tt 1} and {\tt 2} are pushed and subsequently
popped in the same order violates the atomic stack ADT. Yet, while replacing
both values {\tt 1} and {\tt 2} with the value {\tt 1} results in a
\emph{valid} stack execution, replacing them with {\tt 3} and {\tt 4} results
in a violation.

Our algebraic insight is based on grouping the operations of an execution into
\emph{matchings}. Intuitively, operations which refer to the same instances of
values belong to the same matching. For example, a {\tt pop} operation
returning the value {\tt 1} matches a preceding {\tt push(1)} operation. By
comparing executions by the characteristics of their matchings, rather than the
actual data values they use, we capture the relation among data values
independently of the data values themselves. Furthermore, the notion of
matchings provides a key algebraic property of ADTs: the executions of
naturally-occurring concurrent object ADTs are closed under the removal of
matchings. For example, any execution of the atomic stack ADT using values {\tt
1}, {\tt 2}, and {\tt 3} would remain a valid execution were all operations
using the value {\tt 2} deleted. Conversely, any execution which extends a
violating execution with additional matchings is itself a violation. This
property, along with analogous algebraic properties concerning operation order
and completion, allow us to compare executions via a violation-preserving
embedding relation. This relation is a well-founded partial order on
executions, and thus allows us to characterize the infinite set of violations
to an ADT with a finite basis set, ultimately leading to a finite symbolic
representation.

@@

@@

@@

  \item Identification of algebraic properties allowing finite characterization
  and computability of infinite ADT violations sets (§\ref{sec:algorithm}).

  \item An algorithm to infer the matching functions required for our algebraic
  characterization of ADTs (§\ref{sec:matching}).

  \item Empirical validation that our ADT-inference approximation is complete
  in practice, and effective for the detection of observational-refinement
  violations~(§\ref{sec:empirical}).

\end{itemize}
We conclude with a discussion of related work (§\ref{sec:related}).

As far as we are aware, this work is the first to suggest the automatic
generation of symbolic ADT representations for concurrent objects. By removing
the burden of writing formal ADT specifications manually, this work broadens
the scope of efficient symbolic refinement-checking algorithms to
newly-designed ADTs, apart from those few traditionally studied in the
literature.

>>>>

The fundamental challenge is in identifying the algebraic properties of ADTs
which allow us to characterize an infinite set of violating executions with a
finite set of patterns. This characterization is non-trivial since an execution
with more operations than a given violating execution is not necessarily a
violation itself. For instance, an execution which contains only a single
pop operation returning the value $1$ is a violation to the atomic stack
ADT, whereas an execution containing an additional push$(1)$ operation,
overlapping in time with the pop, is not. Further complication arises
from the infinite set of possible data values, i.e.,~method argument and return
values. Our patterns must be sensitive to the \emph{relation} among data values
without being sensitive to the data values themselves. For instance, a
sequential execution in which $1$ and $2$ are pushed and subsequently
popped in the same order violates the atomic stack ADT. Yet, while replacing
both values $1$ and $2$ with the value $1$ results in a
\emph{valid} stack execution, replacing them with $3$ and $4$ results
in a violation.

Our algebraic insight is based on grouping the operations of an execution into
\emph{matchings}. Intuitively, operations which refer to the same instances of
values belong to the same matching. For example, a pop operation
returning the value $1$ matches a preceding push$(1)$ operation. By
comparing executions by the characteristics of their matchings, rather than the
actual data values they use, we capture the relation among data values
independently of the data values themselves. Furthermore, the notion of
matchings provides a key algebraic property of ADTs: the executions of
naturally-occurring concurrent object ADTs are closed under the removal of
matchings. For example, any execution of the atomic stack ADT using values
$1$, $2$, and $3$ would remain a valid execution were all operations
using the value $2$ deleted. Conversely, any execution which extends a
violating execution with additional matchings is itself a violation. This
property, along with analogous algebraic properties concerning operation order
and completion, allow us to compare executions via a violation-preserving
embedding relation. This relation is a well-founded partial order on
executions, and thus allows us to characterize the infinite set of violations
to an ADT with a finite basis set, ultimately leading to a finite symbolic
representation.

@@

Although our approach does require annotating the operations of concurrent
object executions with a matching relation, and we demonstrate that these
relations can be given for naturally-occurring ADTs, we also provide an
automatic means for computing such \emph{matching schemes}. Again by sampling
executions, we leverage automated symbolic reasoning engines to synthesize
matching schemes for which given implementations are closed under the removal
of matches. This further lowers the burden of automated verification. Rather
than providing formal ADT specifications, or even matching schemes, users need
only provide the predicates relevant in the logic of matching schemes, and we
could automatically compute effective matching schemes, and ultimately,
effective symbolic ADT representations.

@@

  \item An abstract notion of execution histories based on groups of matching
  operations (§\ref{sec:histories}).

@@

  \item Identification of the algebraic properties allowing finite
  characterization and computability of infinite ADT violations sets
  (§\ref{sec:patterns}).

  \item The computability of symbolic ADT representations (§\ref{sec:formula}).

  \item An algorithm to infer the matching schemes required for our algebraic
  characterization of ADTs (§\ref{sec:matching}).

  \item An empirical study validating that naturally-occurring ADTs satisfy the
  properties required for completeness of our inference algorithm, and that our
  algorithm computes precise symbolic representations thereof
  (§\ref{sec:nature}).

\end{itemize}
We conclude with a discussion of the limitations of our approach
(§\ref{sec:discussion}) and related work (§\ref{sec:related}).

To the best of our knowledge, this work is the first to suggest the automatic
generation of symbolic ADT representations for concurrent objects. By removing
the burden of writing formal ADT specifications manually, this work broadens
the scope of efficient symbolic refinement-checking algorithms to
newly-designed ADTs, apart from those few traditionally studied in the
literature.
